Generating Random Logic Programs Using Constraint Programming

### 2019-12-30 paper/paper.tex
Counting: Attempt 2

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We also use Choco \cite{choco}. This works with both Prolog
\cite{DBLP:books/daglib/0041598} and ProbLog \cite{DBLP:conf/ijcai/RaedtKT07}.
Tested with SWI-Prolog \cite{DBLP:journals/tplp/WielemakerSTL12}.

Inference options to explore. Logspace vs normal space. Symbolic vs
non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
fsdd, fbdd.

@@

\begin{itemize}
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item A constraint for logical equivalence.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\end{itemize}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses - 1$, let $p_i =
  \variable{headsOfClauses}[i].\variable{predicate}$. Then
  \[
    \variable{sort}((p_i)_{i=0}^{\maxNumClauses - 1}, (p_i)_{i=0}^{\maxNumClauses
      - 1}),
  \]
  i.e., all clauses describing the same predicate are listed together and
  sorted.
\end{constraint}

@@

\begin{constraint}
  $\variable{sort}(\variable{treeStructure}, \variable{treeStructure})$, i.e.,
  \variable{treeStructure} is sorted.
\end{constraint}

@@

Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v=1}^a \binom{a}{v} |\constants{}|^{a-v}
  \sum_{k=0}^{|\variables{}|-1} \binom{v-1}{k}.
\]
First, select the $v$ positions dedicated for variables. The remaining $a-v$
constants can then be filled in $|\constants{}|^{a-v}$ ways. Filling $v$
positions with $k+1$ variables in a non-decreasing manner (without skipping any
variables) can be seen as putting $k$ `bars' in the $v-1$ spaces between $v$
positions. Each bar represents switching to the next variable.
% TODO: the equation is not always defined

Number of programs:
\[
  \left( C^{|\predicates{}|} \prod_{\mathsf{P}/a \in \predicates{}} H_a \right)
  \left( \sum_{n=0}^{\maxNumClauses{} - |\predicates{}|} \left( C
    \sum_{a \in \arities{}} p_aH_a \right)^n \right)
\]
The first half of the equation represents setting up a single clause for every
predicate, i.e., we select its body and its head. The second half iterates over
the number of clauses that could be added. For each of the $n$ additional
clauses, we choose its body and its head (defined by a predicate and its arity).

Parameters that affect this:
\begin{itemize}
\item $\arities{}$ and $(p_a)_{a \in \arities{}}$. Consider even zero arity.
\item $|\predicates{}| = \sum_{a \in \arities{}} p_a$
\item $|\variables{}|$
\item $|\constants{}|$. Consider no variables and no constants.
\item $\maxNumNodes{} \ge 1$
\item $\maxNumClauses{} \ge |\predicates{}|$
\end{itemize}
% TODO: this is without any kind of cycle detection and without probabilities

>>>>

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We also use Choco~4.10.2 \cite{choco}. This works with both Prolog
\cite{DBLP:books/daglib/0041598} and ProbLog \cite{DBLP:conf/ijcai/RaedtKT07}.
Tested with SWI-Prolog \cite{DBLP:journals/tplp/WielemakerSTL12}.

@@

\begin{itemize}
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item A constraint for logical equivalence.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses - 1$, let $p_i =
  \variable{headsOfClauses}[i].\variable{predicate}$. Then
  $(p_i)_{i=0}^{\maxNumClauses - 1}$ is sorted.
\end{constraint}

@@

\begin{constraint}
  \variable{treeStructure} is sorted.
\end{constraint}

@@

Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v=1}^a \binom{a}{v} |\constants{}|^{a-v}
  \sum_{k=0}^{|\variables{}|-1} \binom{v-1}{k}.
\]
First, select the $v$ positions dedicated for variables. The remaining $a-v$
constants can then be filled in $|\constants{}|^{a-v}$ ways. Filling $v$
positions with $k+1$ variables in a non-decreasing manner (without skipping any
variables) can be seen as putting $k$ `bars' in the $v-1$ spaces between $v$
positions. Each bar represents switching to the next variable.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
%      h_1 + \dots + h_{|\predicates{}|} = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \binom{C}{h_i} H_{a_i}^{h_i}
\]

Parameters that affect this:
\begin{itemize}
\item $\arities{}$ and $(p_a)_{a \in \arities{}}$. Consider even zero arity.
\item $|\predicates{}| = \sum_{a \in \arities{}} p_a$
\item $|\variables{}|$
\item $|\constants{}|$. Consider no variables and no constants.
\item $\maxNumNodes{} \ge 1$
\item $\maxNumClauses{} \ge |\predicates{}|$
\end{itemize}
% TODO: this is without any kind of cycle detection and without probabilities
% describe the parameters of the test cases
